---
title: "Application to flight data"
output:
    bookdown::html_document2:
        fig_width: 5
        fig_height: 5
        number_sections: true
        toc: true
        toc_depth: 1
pkgdown:
    as_is: true
bibliography: ../inst/REFERENCES.bib
vignette: >
    %\VignetteIndexEntry{Application to flight data}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

<!-- Rmarkdown requries 4 spaces of indentation -->

# The dataset

```{r, echo = TRUE, message=FALSE, warning=FALSE}
library(graphicalExtremes)
library(ggplot2)
library(igraph)
```

```{r, include=FALSE, eval=FALSE}
rightAlign <- function(txt){
    paste0('<div style="text-align: right">', txt, '</div>')
}
now <- Sys.time()
knitr::knit_hooks$set(timeit = function(before, options, envir) {
    if (before) {
        now <<- Sys.time()
    } else {
        rightAlign(sprintf("\n_Code execution time: %.3f seconds._\n", Sys.time() - now)) 
    }
})
```

```{r, include=FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = '#>',
    error = TRUE,
    fig.align = 'center',
    timeit = TRUE
)

theme_set(
    theme_bw()
    + theme(
        plot.background = element_blank(),
        legend.background = element_blank(),
        strip.background = element_rect(fill = "white"),
        plot.caption=element_text( size=7.5, hjust=0, margin=margin(t=15)),
        text = element_text(size = 11),
        axis.ticks = element_blank(),
        panel.grid.major = element_line(size = 0.25)
    )
)

```

The `flights` dataset contains daily total delays of major U.S. airlines.
For details, see the corresponding documentation page.
Below, we plot all airports in the dataset.

```{r, fig.width=10}
plotFlights(plotConnections = FALSE, map = "world", xyRatio = 2)
```

To perform an example application,
we consider the years 2010 - 2013 and choose (somewhat arbitrarily)
airports from the westcoast with a large number of flights per year.
A similar analysis is perform in @hen2022.
For more detailed explanations see Vignette "applicationDanube" and the help pages of applied functions.
*Note: In the CRAN version of this package, these are the only available years.*
*A version of this package with a larger dataset is available on GitHub*
```{r, fig.width=10}
# Specify years
yearNames <- as.character(seq(2010, 2013))

# Compute departures + arrivals per airport
flightsPerConnection <- apply(flights$flightCounts[, , yearNames], c(1, 2), sum)
flightsPerAirport <- rowSums(flightsPerConnection) + colSums(flightsPerConnection)

# Select airports (more than 1000 flights/year and rough westcoast coordinates)
indKeep <- (
    flightsPerAirport >= length(yearNames) * 1000
    & flights$airports$Longitude < -119 & flights$airports$Longitude > -130
    & flights$airports$Latitude > 25 & flights$airports$Latitude < 50
)
IATAs <- flights$airports$IATA[indKeep]

# Plot airports + connections (indicating number of flights by thickness)
plotFlights(
    IATAs,
    useAirportNFlights = TRUE,
    useConnectionNFlights = TRUE
)
```

Next, we create the matrix containing (departing) flight delays for the considered airports and dates.
```{r}
dates <- as.Date(dimnames(flights$delays)[[1]])
indDates <- format(dates, "%Y") %in% yearNames
mat <- flights$delays[indDates, IATAs, "departures"]
# We remove all rows containing NAs
rowHasNA <- apply(is.na(mat), 1, any)
mat <- mat[!rowHasNA, ]

# Percentile used as threshold when transforming to multivariate Pareto distribution
p <- 0.7

# Empirical extremal correlation matrix
emp_chi_mat <- emp_chi(mat, p = p)

```

This function will be used later to compare fitted parameters to empirical ones.
```{r}
# Utility function to plot fitted parameters against true/empirical ones
plot_fitted_params <- function(G0, G1, xlab = 'Empirical', ylab = 'Fitted'){
    return(
        ggplot()
        + geom_point(aes(
            x = G0[upper.tri(G0)],
            y = G1[upper.tri(G1)]
        ))
        + geom_abline(slope = 1, intercept = 0)
        + xlab(xlab)
        + ylab(ylab)
    )
}
```

# Fitting a model to the flight graph

As a baseline for graphical modelling, we consider the graph with edges
representing at least monthly connections between airports.
```{r}
# Compute undirected flights per connection between selected airports
flightsPerConnectionUD <- flightsPerConnection + t(flightsPerConnection)
flightsPerConnectionUD <- flightsPerConnectionUD[IATAs, IATAs]

# Make flight graph from adjacency matrix
A <- 1 * (flightsPerConnectionUD > length(yearNames) * 12)
flight_graph <- graph_from_adjacency_matrix(A, diag = FALSE, mode = "undirected")

# Plot flight graph
plotFlights(IATAs, graph = flight_graph, clipMap = 1.3, xyRatio = 1)
```

Given the `flight_graph` object, we fit a Hüsler--Reiss model with that graphical structure.
```{r, message=FALSE, warning=FALSE}
# Fit the model
model_fit <- fmpareto_graph_HR(
    data = mat,
    graph = flight_graph,
    p = p,
    method = "vario"
)

# Compute likelihood/ICs
flights_loglik_graph <- loglik_HR(
    data = mat,
    p = p,
    graph = flight_graph,
    Gamma = model_fit
)
cat("Flight graph BIC =", round(flights_loglik_graph['bic'], 2), "\n")

# Plot fitted parameters
plot_fitted_params(emp_chi_mat, Gamma2chi(model_fit))
```


# Fitting a tree model
Next, we fit an extremal tree model to the flight delays using `emst()`.
```{r, collapse=TRUE, fig.align='', fig.show='hold', out.width='50%'}
# Fit the model
flights_emst_fit <- emst(data = mat, p = p, method = "vario")

# Compute likelihood/ICs
flights_loglik_tree <- loglik_HR(
    data = mat,
    p = p,
    Gamma = flights_emst_fit$Gamma,
    graph = flights_emst_fit$graph
)
cat("Tree BIC =", round(flights_loglik_tree['bic'], 2), "\n")

# Plot fitted graph, parameters
plotFlights(
    IATAs,
    graph = flights_emst_fit$graph,
    xyRatio = 1,
    clipMap = 1.3
)
plot_fitted_params(emp_chi_mat, Gamma2chi(flights_emst_fit$Gamma))
```


# Fitting a general model

Lastly, we fit a general graphical model with `eglearn()`, using a suitable list of penalization parameters.
```{r, message=FALSE, warning=FALSE}
# Fit the model
rholist <- seq(0, 0.50, length.out = 11)
flights_eglearn_fit <- eglearn(mat, p = p, rholist = rholist, complete_Gamma = TRUE)

# Compute likelihood/ICs
flights_loglik <- sapply(seq_along(rholist), FUN = function(j) {
    loglik_HR(
        data = mat,
        p = p,
        Gamma = flights_eglearn_fit$Gamma[[j]],
        graph = flights_eglearn_fit$graph[[j]]
    )
})
```

The "best" penalization parameter can be chosen e.g. such that the BIC is minimized.
```{r}
ggplot(
    mapping = aes(x = rholist, y = flights_loglik['bic', ])) +
    geom_line() +
    geom_point(shape = 21, size = 3, stroke = 1, fill = "white") +
    geom_hline(aes(yintercept = flights_loglik_graph['bic']), lty = "dashed") +
    geom_hline(aes(yintercept = flights_loglik_tree['bic']), lty = "dotted") +
    xlab("rho") +
    ylab("BIC") +
    scale_x_continuous(
        breaks = rholist,
        labels = round(rholist, 3),
        sec.axis = sec_axis(
            trans = ~., breaks = rholist,
            labels = sapply(
                flights_eglearn_fit$graph,
                igraph::gsize
            ),
            name = "Number of edges"
        )
)

best_index <- which.min(flights_loglik['bic',])
cat('Best rho =', rholist[best_index], '\n')
cat('Corresponding BIC =', flights_loglik['bic', best_index])
```

We plot the estimated graph and parameters of the best fitted model.
```{r, fig.align='', fig.show='hold', out.width='50%'}
best_graph <- flights_eglearn_fit$graph[[best_index]]
best_Gamma <- flights_eglearn_fit$Gamma[[best_index]]
plotFlights(IATAs, graph = best_graph, clipMap = 1.3, xyRatio = 1)
plot_fitted_params(emp_chi_mat, Gamma2chi(best_Gamma))
```

# References
