---
title: "Exercises"
output:
  bookdown::html_document2:
    fig_width: 5
    fig_height: 5
    number_sections: false
bibliography: ../inst/REFERENCES.bib
toc:
  depth: 2
pkgdown:
  as_is: true 
vignette: >
  %\VignetteIndexEntry{Exercises}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
  
  \DeclareMathOperator{\Var}{Var}
  \newcommand{\g}[1]{\mathbf{#1}}
  \DeclareMathOperator{\MST}{mst}
  \DeclareMathOperator{\argmin}{arg\,min}
  
  ```{r, include = FALSE}
knitr::opts_chunk$set(
collapse = TRUE,
comment = ">"
)
```

```{r setup, include=FALSE}
library(graphicalExtremes)
library(dplyr)
library(ggplot2)
theme_set(theme_bw() +
            theme(plot.background = element_blank(),
                  legend.background = element_blank(),
                  strip.background = element_rect(fill = "white"),
                  plot.caption=element_text(size=7.5, hjust=0, 
                                            margin=margin(t=15)),
                  text = element_text(size = 11),
                  axis.ticks = element_blank(),
                  panel.grid.major = element_line(size = 0.25)))
```


The `flights()`dataset contains daily total delays (Jan 1, 2015 -- Dec 31, 2015) of Southwest Airlines at different airports in California, Nevada, Arizona, Utah, and Texas.
The object `flights()` is a list made of:

* `data`: a `dplyr::tibble()`  containing the delay data. Each row corresponds to a day and each column to an airport.
* `airports`: a `dplyr::tibble()` containing the details for each airport.
* `connections`: a `dplyr::tibble()` containing the details of the _undirected_ flights.

In the code below, we show how to load the dataset and plot the flight connections.

```{r, fig.align='center', fig.width=10}
# Load the dataset
data <- flights$data
airports <- flights$airports

airports_id <- tibble(ORIGIN_AIRPORT = airports$ORIGIN_AIRPORT, 
                      ID = seq_len(nrow(airports)))
connections <- flights$connections %>% 
  left_join(airports_id, by = c("ORIGIN_AIRPORT" = "ORIGIN_AIRPORT")) %>% 
  left_join(airports_id, by = c("DESTINATION_AIRPORT" = "ORIGIN_AIRPORT"), 
            suffix = c(".origin", ".dest"))

mat <- data %>% 
  select(-DATE) %>% 
  as.matrix()

# Plot flight connection
ggplot() +
  geom_polygon(data = map_data("usa"),
               aes(x = long, y = lat, group = group), color = "grey65",
               fill = "#f9f9f9", size = 0.2) +
  geom_point(data = airports,
             aes(x = LONGITUDE, y = LATITUDE, size = N_FLIGHTS),
             alpha = 1) +
  geom_curve(data = connections,
             aes(x = LONGITUDE.origin, xend = LONGITUDE.dest,
                 y = LATITUDE.origin, yend = LATITUDE.dest,
                 size = N_FLIGHTS),
             alpha = .2, curvature = 0)

```

Also, it is useful to create an `igraph::graph()` object for the flights connections.
```{r}
flight_graph <- connections %>%
  ungroup() %>% 
  select(ID.origin, ID.dest) %>%
  as.matrix() %>%
  igraph::graph_from_edgelist(directed = FALSE)

flight_graph2 <- connections %>% 
  select(ORIGIN_AIRPORT, DESTINATION_AIRPORT) %>%
  as.matrix() %>%
  igraph::graph_from_edgelist(directed = FALSE)


igraph::E(flight_graph)
mat

```


In the exercises below we are modeling the extremal dependence of the flight delays at the 24 airports. Note that the `flight_graph` does not necessarily represent the extremal graphical structure, but it can be used as a comparison to the estimated graphs. 

# Part A
1. Fit an extremal tree model to the flight delays using `emst()`, choosing the threshold `p = 0.7`. 
```{r, collapse=TRUE, fig.align='center'}
p <- .7
flights_emst_fit <- emst(data = mat, p = p, method = "vario")
igraph::V(flights_emst_fit$graph)$name <- colnames(mat)

```

2. Plot the estimated tree on the US map and interpret the results.
```{r, fig.align='center', fig.width=10}
flights_connections_est <- igraph::get.edgelist(flights_emst_fit$graph) %>%
  as_tibble(.name_repair = ~ c("ORIGIN_AIRPORT", "DESTINATION_AIRPORT")) %>%
  left_join(flights$airports, by = c("ORIGIN_AIRPORT" = "ORIGIN_AIRPORT")) %>%
  left_join(flights$airports, by = c("DESTINATION_AIRPORT" = "ORIGIN_AIRPORT"),
            suffix = c(".origin", ".dest")) %>%
  select(ORIGIN_AIRPORT, DESTINATION_AIRPORT,
         LATITUDE.origin, LONGITUDE.origin,
         LATITUDE.dest, LONGITUDE.dest)

ggplot() +
  geom_polygon(data = map_data("usa"),
               aes(x = long, y = lat, group = group), color = "grey65",
               fill = "#f9f9f9", size = 0.2) +
  geom_point(data = airports,
             aes(x = LONGITUDE, y = LATITUDE, size = N_FLIGHTS),
             alpha = 1) +
  geom_curve(data = flights_connections_est,
             aes(x = LONGITUDE.origin, xend = LONGITUDE.dest,
                 y = LATITUDE.origin, yend = LATITUDE.dest),
             alpha = .4, curvature = 0)

```

3. Compute the BIC value of the fitted tree model.
```{r}
flights_loglik_tree <- loglik_HR(data=mat, p=p,
                                 Gamma = flights_emst_fit$Gamma,
                                 graph = flights_emst_fit$graph)

paste0("Tree BIC = ", flights_loglik_tree[3] %>% round(2))

```

4. Plot the empirical $\chi$ coefficient against the $\chi$ coefficient implied by the fitted model.

```{r, fig.align='center'}
emp_chi_mat <- emp_chi(mat, p = p)

ggplot() +
  geom_point(aes(x = c(Gamma2chi(flights_emst_fit$Gamma)),
                 y = c(emp_chi_mat))) +
  geom_abline(slope = 1, intercept = 0) +
  xlab("Fitted") +
  ylab("Empirical")
```

5. Given the `flight_graph` object, fit a HR graphical model using `fmpareto_graph_HR()`.
```{r}
model_fit <- fmpareto_graph_HR(data = mat,
                               graph = flight_graph, p = p, method = "vario")

compl_gamma <- complete_Gamma(emp_vario(data = mat, p = .7), graph = flight_graph2)
outliers <- which(abs(Gamma2chi(compl_gamma) - 
            Gamma2chi(emp_vario(data = mat, p = .7))) > .05,
      arr.ind = TRUE) %>% 
  as_tibble() %>% 
  rename(ID.origin = row, ID.dest = col) %>% 
  left_join(airports_id, by = c("ID.origin" = "ID")) %>% 
  left_join(airports_id, by = c("ID.dest" = "ID")) %>% 
  mutate(name1 = paste0(ORIGIN_AIRPORT.x, ORIGIN_AIRPORT.y),
         name2 = paste0(ORIGIN_AIRPORT.y, ORIGIN_AIRPORT.x))


unique(c(outliers$name1, outliers$name2)) %in% 
  paste0(connections$ORIGIN_AIRPORT, connections$DESTINATION_AIRPORT)

compl_gamma2 <- complete_Gamma(emp_vario(data = mat, p = .8), graph = flight_graph)
```

6. Compute the BIC value for the `flight_graph` object and the corresponding `Gamma` matrix obtained at Step 5.

```{r}
flights_loglik_graph <- loglik_HR(data = mat, 
                                  p = p, graph = flight_graph,
                                  Gamma = model_fit$Gamma)
paste0("Tree BIC = ", flights_loglik_graph[3] %>% round(2))

```

7. Plot the empirical $\chi$ coefficient against the $\chi$ coefficient implied by the fitted model of Step 5.

```{r, fig.align='center'}
ggplot() +
  geom_point(aes(x = c(Gamma2chi(model_fit$Gamma)),
                 y = c(Gamma2chi(emp_vario(mat, p = p))))) +
  geom_abline(slope = 1, intercept = 0) +
  xlab("Fitted") +
  ylab("Empirical")


Gamma2chi(model_fit$Gamma) == Gamma2chi(emp_vario(mat, p = p))

outlier_inds <- which(upper.tri(Gamma2chi(emp_vario(mat, p = p))) & 
                        Gamma2chi(model_fit$Gamma) < 0.4 &
                        Gamma2chi(emp_vario(mat, p = p)) > 0.5,
                      arr.ind = TRUE)

```

7. Do you observe any "outliers" in the previous plot? Can you provide an explanation for these?
```{r, fig.align='center', fig.width=10}
# Select suspicious flight connections
outlier_inds2 <- which(upper.tri(emp_chi_mat) & 
                        Gamma2chi(model_fit$Gamma) < 0.4 &
                        emp_chi_mat > 0.5,
                      arr.ind = TRUE)

outlier_connections <- tibble(
  ORIGIN_AIRPORT = colnames(mat)[outlier_inds[, 1]],
  DESTINATION_AIRPORT = colnames(mat)[outlier_inds[, 2]]
) %>% 
  left_join(airports, by = c("ORIGIN_AIRPORT" = "ORIGIN_AIRPORT")) %>%
  left_join(airports, by = c("DESTINATION_AIRPORT" = "ORIGIN_AIRPORT"),
            suffix = c(".origin", ".dest")) %>%
  select(ORIGIN_AIRPORT, DESTINATION_AIRPORT,
         LATITUDE.origin, LONGITUDE.origin,
         LATITUDE.dest, LONGITUDE.dest)


# Plot San Diego International Airport delays
ggplot(data) +
  geom_point(aes(x = DATE, y = SAN))

# Possible explanation July 18, 2015 from dataset  (https://www.sandiegouniontribune.com/sdut-thunderstorms-fire-dolors-2015jul18-story.html)
# mat_wo_storm <- data %>% 
#   filter(lubridate::month(DATE) != 7) %>%
#   select(-DATE) %>% 
#   as.matrix()
# 
# # Fit again the model
# model_fit_wo_storm <- fmpareto_graph_HR(data = mat_wo_storm,
#                                graph = flight_graph, p = p, method = "vario")
# 
# # Plot goodness of fit
# ggplot() +
#   geom_point(aes(x = c(Gamma2chi(model_fit_wo_storm$Gamma)),
#                  y = c(emp_chi(mat_wo_storm, p = p)))) +
#   geom_abline(slope = 1, intercept = 0) +
#   xlab("Fitted") +
#   ylab("Empirical")

```


# Part B

1. Fit an extremal graphical lasso model with `eglasso()`, choosing threshold `p = 0.7`, and `rholist = seq(1e-4, 0.10, length.out = 10)`.

```{r}
Gamma <- emp_vario(mat, p = p)
rholist = seq(1e-4, 0.10, length.out = 10)
flights_eglasso_fit <- eglasso(Gamma, rholist = rholist, complete_Gamma = TRUE)

# Add labels to graph edges
flights_eglasso_fit$graph <- lapply(seq_along(rholist), function(i) {
  est_graph <- flights_eglasso_fit$graph[[i]]
  igraph::V(est_graph)$name <- colnames(mat)
  est_graph
})
```

2. Plot the estimated graph on the US map for different values of `rho` and interpret the results.

```{r, fig.align='center', fig.width=10}
flights_connections_est <-
  igraph::get.edgelist(flights_eglasso_fit$graph[[10]]) %>%
  as_tibble(.name_repair = ~ c("ORIGIN_AIRPORT", "DESTINATION_AIRPORT")) %>%
  left_join(flights$airports, by = c("ORIGIN_AIRPORT" = "ORIGIN_AIRPORT")) %>%
  left_join(flights$airports, by = c("DESTINATION_AIRPORT" = "ORIGIN_AIRPORT"),
            suffix = c(".origin", ".dest")) %>%
  select(ORIGIN_AIRPORT, DESTINATION_AIRPORT,
         LATITUDE.origin, LONGITUDE.origin,
         LATITUDE.dest, LONGITUDE.dest)

ggplot() +
  geom_polygon(data = map_data("usa"),
               aes(x = long, y = lat, group = group), color = "grey65",
               fill = "#f9f9f9", size = 0.2) +
  geom_point(data = airports,
             aes(x = LONGITUDE, y = LATITUDE, size = N_FLIGHTS),
             alpha = 1) +
  geom_curve(data = flights_connections_est,
             aes(x = LONGITUDE.origin, xend = LONGITUDE.dest,
                 y = LATITUDE.origin, yend = LATITUDE.dest),
             alpha = .4, curvature = 0)
```

3. Compute and plot the BIC values of the estimated models for different values of `rho`.

```{r, fig.align='center'}
flights_loglik <- sapply(seq_along(rholist), FUN = function(j) {
 loglik_HR(data=mat, p=p, 
           Gamma = flights_eglasso_fit$Gamma[[j]],
           graph = flights_eglasso_fit$graph[[j]] )
})

ggplot(mapping = aes(x = rholist, y = flights_loglik[3, ])) +
  geom_line() +
  geom_point(shape = 21, size = 3, stroke = 1, fill = "white") +
  geom_hline(aes(yintercept = flights_loglik_tree[3]), lty = "dashed") +
  xlab("rho") +
  ylab("BIC") +
  scale_x_continuous(
    breaks = rholist,
    labels = round(rholist, 3),
    sec.axis = sec_axis(trans=~., breaks = rholist,
                        labels = sapply(flights_eglasso_fit$graph, 
                                        igraph::gsize),
                        name="Number of edges")
  )

```

<!-- 3b. Another way to choose best `rho` is to compute `loglik` on test data (!!! if you have more data). -->

4. Plot the empirical $\chi$ coefficient against the $\chi$ coefficient implied by the *best* fitted model.

```{r, fig.align='center'}
best_Gamma <- flights_eglasso_fit$Gamma[[which.min(flights_loglik[3,])]]

ggplot() +
  geom_point(aes(x = c(Gamma2chi(best_Gamma)),
                 y = c(emp_chi_mat))) +
  geom_abline(slope = 1, intercept = 0) +
  xlab("Fitted") +
  ylab("Empirical")

```

5. The function `eglasso()` accepts as input the variogram matrix `Gamma`. Try to estimate the variogram matrix using different methods, such as:

* `emp_vario()`, fixing some `k`, e.g., `k = 1`.
* `emp_chi()` and then transforming the result with `chi2Gamma()`.
* `graphicalExtremes:::ml_weight_matrix()`, which is an internal function.

With these different estimates for the variogram, fit an `eglasso()` model and plot the different BIC curves.

_Hint_: For some of the methods, you might need to adjust the `rholist` provided to `eglasso()`.

```{r, fig.align='center'}
Gamma_vario_k_1 <- emp_vario(data = mat, k = 1, p = p)

flights_eglasso_fit <- eglasso(Gamma_vario_k_1, rholist, complete_Gamma = TRUE)

flights_loglik <- sapply(seq_along(rholist), FUN = function(j) {
 loglik_HR(data=mat, p=p, 
           Gamma = flights_eglasso_fit$Gamma[[j]],
           graph = flights_eglasso_fit$graph[[j]] )
})

ggplot(mapping = aes(x = rholist, y = flights_loglik[3, ])) +
  geom_line() +
  geom_point(shape = 21, size = 3, stroke = 1, fill = "white") +
  geom_hline(aes(yintercept = flights_loglik_tree[3]), lty = "dashed") +
  xlab("rho") +
  ylab("BIC") +
  scale_x_continuous(
    breaks = rholist,
    labels = round(rholist, 3),
    sec.axis = sec_axis(trans=~., breaks = rholist,
                        labels = sapply(flights_eglasso_fit$graph, 
                                        igraph::gsize),
                        name="Number of edges")
  )

```


```{r, fig.align='center'}
Gamma_from_chi <- chi2Gamma(chi = emp_chi(data = mat, p = p))

matrixcalc::is.positive.definite(Gamma2Sigma(Gamma_from_chi, k = 1))

# !!! opposite behaviour wrt rho
rholist <- seq(0.09, 0.2, length.out = 10)
Gamma_ml <- graphicalExtremes:::ml_weight_matrix(data = mat,
                                                 p = p)
flights_eglasso_fit <- eglasso(Gamma_from_chi, rholist, 
                               complete_Gamma = TRUE)

not_na <- !is.na(flights_eglasso_fit$Gamma)

flights_loglik <- sapply(seq_along(rholist[not_na]), FUN = function(j) {
 loglik_HR(data=mat, p=p, 
           Gamma = flights_eglasso_fit$Gamma[[j]],
           graph = flights_eglasso_fit$graph[[j]] )
})

ggplot(mapping = aes(x = rholist[not_na], y = flights_loglik[3, ])) +
  geom_line() +
  geom_point(shape = 21, size = 3, stroke = 1, fill = "white") +
  geom_hline(aes(yintercept = flights_loglik_tree[3]), lty = "dashed") +
  xlab("rho") +
  ylab("BIC") +
  scale_x_continuous(
    breaks = rholist,
    labels = round(rholist, 3),
    sec.axis = sec_axis(trans=~., breaks = rholist,
                        labels = sapply(flights_eglasso_fit$graph, 
                                        igraph::gsize),
                        name="Number of edges")
  )
```

